Binning Strategies and Related Loss for Binned Scatterplots of Large Data

### 2016-12-04 AuthorResponseToReviewers.Rmd

### Responses to Reviewer 1 Comments/Questions

*The authors discuss different binning strategies and the resulting aggregation- induced errors. Both errors, loss of spatial resolution and loss of frequency information are introduced and formalized. Binning is an important preprocessing step and research in this area is highly relevant. However, there are some major points missing, making this work less convincing.*

*The most crucial point is in my opinion the missing discussion of the influence of different data distributions. In the visualization community exists research on different descriptors for two-dimensional data called Scagnostics. When I started reading this article, I expected a thorough discussion of different data distributions/properties and the influence on different binning strategies and error outcomes.*

- We are very grateful that the review brought the literature on scagnostics to our attention; the work and scagnostic metrics formalized in the works by Wilkerson, Anand, Grosmann and Wills will be valuable to the discussion of binned scatterplots. Specifically scagnostic literature contributes to Section 5 of our paper in the exploration of loss under a variety of simulated distribution. You will find that a discussion of the relevance of scagnostics has been added immediately preceding the start of subsection 5.1. It is interesting that the striation score is so small, given that the data resolution issue can lead to artificially induced stripping is seen in so prominently in Figure 6.

@@

*In general, I think this paper addresses an important topic. Guidelines on choosing binning algorithms will be immensely helpful for practitioners dealing with big data. I am, however, reserved in recommending acceptance because of two main concerns: lack of novelty and unconvincing results.*

*Lack of Novelty:*
*Section 2 and 3 do a good job in reviewing the concepts of binned aggregation and varieties of binning algorithms. However these are fairly well known and the discussion does not seem to contribute novel ideas. The only thing that I haven’t seen before is the random binning approach. Is this a new idea proposed by you? If so, please state clearly. I also wonder how and why random binning can be useful. Please provide scenarios in which it would make sense to use random binning.*

*For the other ideas covered in these two sections, please properly cite relevant work.*

- You are correct that many of the ideas in Sections 2 and 3 are not original, with the exception of the random binning approach, where the methodologies for random-binned aggregation and post processing for net spatial loss are original to our best knowledge. The goal of these sections is to bring many binning related concepts under a unified notation for the benefit of clarity in the sections to follow. For the random binning, clearer language of our idea and a statement has been added to address the motivation for its introduction: deterministic one-sided interval bins lead to lopsided allocation of points on bin boundaries, whereas a random assignment is expected to split the points evenly across bins. For the unoriginal ideas, additional citations have been made to place the unifying notation better within the existing literature.

@@

*In visualization practices, it is common to use both ordinal scales as well as continuous scales when mapping numbers to colors. How might you address the issue of frequency loss when continuous scales are used?*

- The need for future perception research to quantify the implicit frequency loss when using continuous shading scales is now noted in the discussion section of the paper.

*I don’t buy the advice to “use between four and seven frequency”. There simply isn’t empirical evidence in the paper to support this claim.*

- We do not want to allow more color bin than we can *easily* differentiate because the our expressed purpose for working with discretized color scales in this scenario is to have made loss of frequency information *entirely* explicit. While working with discretizing shades we found ourselves worrying about our ability to perfectly match bins to a legend after approximately 6 shades. We do not claim to have gathered empirical evidence, rather we defer to Healey and Enn's Theory on discernible colors (typically 4-7) to back out experience. Figure 9 was included for the reader to assess if they felt our suggestion was justifiable. A statement clarifying that the recommendations is to bolster the reliable perceptual mapping has been added to the discussion section.

### 2016-12-04 versionForJCGS/PaperJCGS.Rnw
Spell Check and minor edit

An alternative to the rectangular binning process, is a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
%
for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers. This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers. \ktm{The impetus for developing the random binning assignment was to soften the hard breaks associated with one-sided interval assignment for scenarios where many points fall directly on bin boundaries by allowing points to be divided stochastically.} The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.

>>>>

As an alternative to the rectangular binning process, \ktm{We propose} a \textit{random binning} algorithm which utilizes a non-deterministic bin function $b^r_X(\cdot)$ to randomly assigns an observation, $x_i$, to a bin center, $x^\ast$, from a set of possible bins. In this paper, we will consider the simplest case of just two bins, so that without loss of generality we can assume that $x_i$ lies between bin centers $x^\ast_j$ and $x^\ast_{j+1}$. The bin function assigns $x_i$ to a bin center with a probability inversely proportional to the distance to that bin center; the closer a value is to a bin center, the higher the probability the value is assigned to that bin center. More formally,
%
\begin{eqnarray}\label{randbin1}
b^r_X(x_i) = \left\{\begin{array}{ll} 
  x^\ast_j & \text{ with probability } (x^\ast_{j+1} - x_i)/(x^\ast_{j+1} -x^\ast_{j}) \\
  x^\ast_{j+1} &\text{ with probability }  (x_i - x^\ast_{j})/(x^\ast_{j+1} -x^\ast_{j})
  \end{array}\right.
\end{eqnarray}  
%
for $x_i \in [x^\ast_{j+1}, x^\ast_{j}]$. In Table~\ref{tab:rectbinning} we note that this random binning algorithm does not specify bin boundaries; only a sequence of bin centers. This method is easily extensible to also map $x_i$ into more than two bins and can accommodate non-uniform distribution of bin centers. \ktm{The impetus for developing the random binning assignment was to soften the hard breaks associated with one-sided interval assignment for scenarios where many points fall directly on bin boundaries by allowing points to be divided stochastically.} The bin boundaries and centers for standard and random rectangular binning algorithms can be found in Table~\ref{tab:rectbinning}.
